`make.modifiers`

### 2023-04-05 inst/gitbook/02_make.traits.Rmd
manual big update

In `dads`, traits are simulating by providing a `traits` object to the `traits` argument.
This object is generated using the `make.traits`.

@@

The function `make.traits` allows you to design the process of a trait or a set of traits.
Here, the process of a trait designates the rules to generate the trait through time while simulating a phylogeny.
This process can depend on the previous state in the tree (i.e. the trait of the ancestor) and the branch length to the descendant.
One classic example is the [Brownian motion process (or Weiner process)](https://en.wikipedia.org/wiki/Brownian_motion).
Note that it _can_ depend on both the ancestor and the branch length but does _not necessary needs_ (i.e. the process can be only based on the previous state or only on branch length or on neither).

@@

Trait processes in `dads` are functions that must always intake the following arguments by default.

 * `x0`: the previous trait value(s)
 * `edge.length`: the branch length value
 * `...`: a placeholder for any extra arguments

For example, the following function would be a valid process that always generate the _true_ trait value: 42!.
In this example, the process is not dependent on neither the previous state (`x0`) and the branch length (`edge.length`).

@@

> Note that in this function definition the arguments `x0` and `edge.length` have a default value set to `0` and `1` respectively. In practice, these arguments are effectively set to the correct values in the `dads` internal function (i.e. whatever `x0` and `edge.length` are at that specific time of the process) but providing a default can help speed up the algorithms (specifically all the internal checks).

@@

```{r}
## A wrongly formated process
invalid.process <- function(x0 = 0) {
    return(rnorm(1, mean = x0))
}
```

@@

You can design your own process as a function (as long as it has a valid syntax).
Alternatively, the `dads` package proposes inbuilt processes, namely a multidimensional Brownian motion (`BM.process`) or a a multidimensional Ornstein-Uhlenbeck process (`OU.process`).
You can find the list of implemented process by looking at the `?trait.process` manual page in `R`.

Once a process is chosen, you can feed it to the `make.traits` function:

@@

This creates `"dads"` `"traits"` objects that you can print, and visualise using the `plot` function:

@@

As traits can get more and more complex, the automatic printing of its summary allows for a easier display of what's in the traits object.

Note that it is possible to make `"traits"` objects with multiple processes (that can be the same):

```{r}
## 4 traits: two BM, one OU and one normal non process
four_traits <- make.traits(process = c(BM.process,
                                       BM.process,
                                       OU.process,
                                       no.process))
four_traits
```

@@

```{r}
## Plot options (4 plots in one window)
par(mfrow = c(2,2))
plot(four_traits, trait = 1)
plot(four_traits, trait = 2)
plot(four_traits, trait = 3)
plot(four_traits, trait = 4)
```

@@

Two further important arguments are `n` the number of traits per process and `start` the starting values for all traits.
By default they are set to `n = 1` and `start = 0`.
This means that `make.traits` will assume that your processes are always unidimensional by default and that they always start with the value `0`.
It is however possible to change these values.

For example you can use the following to create a three dimensional Brownian motion with each dimensions starting with the value `1`:

@@

Note that the number of traits are distributed per processes.
If the traits contains multiple process, the number of traits are distributed per processes:

@@

### What even is a trait?

The definition of what a trait is can vary quiet a lot depending on the context of an analysis and the field.
<!--TODO: add REF e.g.  https://onlinelibrary.wiley.com/doi/full/10.1002/ece3.8321 -->
Because it would be impossible to accommodate all definitions in `dads` we had to go with an arbitrary one: a trait is whatever you define as a trait!
A trait can be uni-dimensional as the measurement of a feature of an organism (e.g. leaf surface, femur length, etc...) but can also be multi-dimensional description of a feature, for example in 3D geometric morphometric, a trait could be defined as "position of landmark X" (which will be a trait with three dimensions, _x_, _y_ and _z_) or in ecology, the location of a plant can be a single 2D trait (expressed as latitude and longitude).
In `dads`, the **process** corresponds to this trait definition (e.g. a process can be of n-dimensions and represents one organisms feature) and the **traits** represents the number of dimensions in total. 
So in the examples above, this is how the following traits are interpreted by `dads`:

@@

You can also feed extra arguments to your process(es) functions. For example, the inbuilt process `no.process` (that is just a number generator not based on the previous value `x0` or the branch length) can intake a specific random number generator as a function:

@@

If one process do not need extra argument you must still give it and extra `NULL` process argument:

@@

You can also add traits to already existing trait objects using the simple `add` option.
This option just intakes a `"dads"` `"traits"` object and the additional process(es) will be added to it. For example:

```{r}
## Creating on simple default Brownian motion
one_process <- make.traits(trait.names = "BM")

@@

`traits` objects also allow a background trait to be used when traits are simulated ([step 3 here](#bdalgorithm)).
This basically allows traits to be simulating for _all_ tips whenever a trait is generated for one tip.
This can be useful for keeping track of trait values along the simulation (_cf_ just at bifurcating nodes).
The `background` argument takes any output from the `make.traits` function in a nested way:

@@

Note that technically you can nest as many background traits as you want (e.g. `make.traits(background = make.traits(background = make.traits(...)))` is valid).
However, you should always make sure that the background trait has the same dimensions as the normal (main) trait.
When using a trait with background, your tree will have internal singleton nodes (i.e. nodes linking to one ancestor and only one descendant).
You can remove these nodes using the [`drop.things` function](#dropthings).

@@

Alternatively to having a background trait, you can also simulate a tree by generating traits at specific time steps with the `save.steps` option in `dads`.
This will apply the `traits` object to all lineages currently alive at the required time steps.
These time steps can be either regular by providing a single numeric value; e.g. `save.steps = 0.1` will get a snapshot of the trait values every 0.1 units of time.
Or specific, by providing a specific set of values; e.g. `save.steps = c(1, 1.2, 3)` will make a snapshot of the trait values at the required time steps.

@@

 * `BM.process`: this is the well famous [Brownian motion process](https://en.wikipedia.org/wiki/Brownian_motion).

@@

 * `no.process`: this process has... no process. In other words, this is a non time dependent process (the simulated value does not depends on the ancestors' value nor the branch length). It's basically a place holder for a random sampling function like `rnorm` (default), `runif`, `rlnorm`, etc...

@@

 * `multi.peak.process`: this process is a modified version of the `OU.process` that can take multiple local optima. The default OU process has one optimum towards which the values are drawn with the alpha parameter (the elastic band) which is usually 0. However, with this `multi.peak.process` we can set multiple values towards which values can be attracted with the same alpha parameter.

@@

 * `repulsion.process`: this is a modified version of the `BM.process` where instead of accumulating normally through time, new trait values are more likely to be different than their ancestor following a `repulsion` parameter.

@@

**This bit is more for development.**
We highly suggest leaving `test = TRUE` so that `make.traits` returns an error if a process or its additional arguments (`process.args`) are not formatted correctly.
`make.traits` will send an error if the trait cannot be directly passed to `dads`.
However, in some specific cases (again, probably mainly for development and debugging) it could be useful to skip the tests using `test = FALSE`.

>>>>

In `dads`, traits are simulating by providing a `traits` object to the `traits` argument.
This object is generated using `make.traits`.

> **What is a trait in biology?** This is one of these epistemological questions like what _is_ a species and is not covered in this package.
For the purpose of _this manual_, a **biological trait** can be any coherent character of any number of dimensions which may or may not be independent of other traits.
Basically anything can be a trait if it's defined by the user (e.g. one trait can be a 20 dimensional Brownian process that's correlated to a 1D OU process - which is another trait).
For the purpose of the syntax in _this package_ a **trait in `dads`** is a column in the output trait matrix.
For example, one might be interested in a biological trait that is "beak shape" which is defined as one complex biological trait that is the Procrustes superimposition of 4 3D landmarks and 300 3D semi-landmarks.
This single "beak shape" **biological trait** is then composed of 912 **traits in `dads`** (i.e. a matrix of $(4+300)\times3$ columns).

@@

The function `make.traits` allows you to design the process of a trait or a set of traits.
Here, the process of a trait designates the rules needed to generate the trait through time while simulating a phylogeny.
This process can depend on the previous state in the tree (i.e. the trait of the ancestor) and the branch length to the descendant.
One classic example is the [Brownian motion process (or Weiner process)](https://en.wikipedia.org/wiki/Brownian_motion).
Note that it _can_ depend on both the ancestor and the branch length but does _not necessarily need to_, i.e. the process can be based only on the previous state or only on branch length or on neither.

@@

Trait processes in `dads` are functions that must always take the following arguments by default.

 * `x0`: the previous trait value(s)
 * `edge.length`: the branch length
 * `...`: a placeholder for any extra arguments

For example, the following function would be a valid process that always generate the _true_ trait value: 42!.
In this example, the process is not dependent on either the previous state (`x0`) or the branch length (`edge.length`).

@@

> Note that in this function definition the arguments `x0` and `edge.length` have a default value of `0` and `1` respectively. In practice, these arguments are effectively set to the correct values in the `dads` internal function (i.e. whatever `x0` and `edge.length` are at that specific time of the process) but providing a default can help speed up the algorithms (specifically all the internal checks).

@@

```{r}
## A wrongly formatted process
invalid.process <- function(x0 = 0) {
    return(rnorm(1, mean = x0))
}
```

This will not work in `make.traits` (see below).

@@

You can design your own process as a function, as long as it has a valid syntax.
Alternatively, the `dads` package has several inbuilt processes, namely a multidimensional Brownian motion (`BM.process`) or a multidimensional Ornstein-Uhlenbeck process (`OU.process`).
You can find the list of implemented processes by looking at the `?trait.process` manual page in `R`.

Once a process is chosen, you can feed it into the `make.traits` function:

@@

This creates `"dads"` `"traits"` objects that you can print and visualise using the `plot` function:

@@

As traits can get more and more complex, the automatic printing of its summary allows for a easier display of what's in the `"traits"` object.

Note that it is possible to make `"traits"` objects with multiple traits and multiple processes (that can be the same or different for each trait):

```{r}
## Four traits: two BM, one OU and one normal non process
four_traits <- make.traits(process = c(BM.process,
                                       BM.process,
                                       OU.process,
                                       no.process))
four_traits
```

@@

```{r}
## Plot options (4 plots in one window)
par(mfrow = c(2,2))
plot(four_traits, trait = 1)
plot(four_traits, trait = 2)
plot(four_traits, trait = 3)
plot(four_traits, trait = 4)
par(mfrow = c(1,1))
```

@@

Two further important arguments are `n` the number of traits per process and `start` the starting values for all traits.
By default they are set to `n = 1` and `start = 0`.
This means that `make.traits` will assume that your processes are always unidimensional by default and that they always start with the value `0`.
It is possible however, to change these values.

For example you can use the following to create a three dimensional Brownian motion with each dimension starting with the value `1`:

@@

Note that the number of traits are distributed per processes.
If the traits contains multiple processes, the number of traits are distributed per processes:

@@

### What is a trait in `dads`?

Because it would be impossible to accommodate all definitions of a trait in `dads` we chose an arbitrary one: a trait is whatever you define as a trait!
A trait can be uni-dimensional as the measurement of a feature of an organism (e.g. leaf surface, femur length, etc.) but can also be a multi-dimensional description of a feature, for example in 3D geometric morphometric context, a trait could be defined as "position of landmark X" (which will be a trait with three dimensions, _x_, _y_ and _z_) or in ecology, the location of a plant can be expressed as latitude and longitude coordinates.
In `dads`, the **process** corresponds to this trait definition (e.g. a process can be of n-dimensions and represents one organisms feature) and the **traits** represents the number of dimensions in total. 
So in the examples above, this is how the following traits are interpreted by `dads`:

@@

You can also feed extra arguments to your process(es) functions. 
For example, the inbuilt process `no.process` (that is just a number generator not based on the previous value `x0` or the branch length) can take a specific random number generator as a function:

@@

If one process does not need extra argument you must still give it an extra `NULL` process argument:

@@

You can also add traits to already existing trait objects using the simple `add` option.
This option just takes a `"dads"` `"traits"` object and the additional process(es) will be added to it. 
For example:

```{r}
## Creating a simple default Brownian motion
one_process <- make.traits(trait.names = "BM")

@@

`traits` objects also allow a background trait to be used when traits are simulated ([step 3 here](#bdalgorithm)).
This allows traits to be simulated for _all_ tips whenever a trait is generated for one tip.
This can be useful for keeping track of trait values along the simulation (_cf_ just at bifurcating nodes).
The `background` argument takes any output from the `make.traits` function in a nested way:

@@

Note that technically you can nest as many background traits as you want (e.g. `make.traits(background = make.traits(background = make.traits(...)))` is valid).
However, you should always make sure that the background trait has the same dimensions as the main trait.
When using a trait with background, your tree will have internal singleton nodes (i.e. nodes linking to one ancestor and only one descendant).
You can remove these nodes using the [`drop.things` function](#dropthings).

@@

You can also simulate a tree by generating traits at specific time steps with the `save.steps` option in `dads`.
This will apply the `traits` object to all lineages currently alive at the required time steps.
These time steps can be either regular by providing a single numeric value; e.g. `save.steps = 0.1` will take a snapshot of the trait values every 0.1 units of time, or specific, by providing a specific set of values; e.g. `save.steps = c(1, 1.2, 3)` will take a snapshot of the trait values at the required time steps.

@@

 * `BM.process`: this is the well known [Brownian motion process](https://en.wikipedia.org/wiki/Brownian_motion).

@@

 * `no.process`: this process has... no process. In other words, this is a non time dependent process; the simulated value does not depends on the ancestors' value nor the branch length. It's basically a place holder for a random sampling function like `rnorm` (default), `runif`, `rlnorm`, etc.

@@

 * `multi.peak.process`: this process is a modified version of the `OU.process` that can take multiple local long-term mean values. The default OU process has one long-term mean towards which the values are drawn with the alpha parameter (the elastic band). The single long-term mean is usually 0. However, with this `multi.peak.process` we can set multiple values towards which values can be attracted with the same alpha parameter.

@@

 * `repulsion.process`: this is a modified version of the `BM.process` where instead of accumulating gradually through time, new trait values are more likely to be different to the traits of their ancestor following a `repulsion` parameter.

@@

**This bit is still in development.**
We highly suggest leaving `test = TRUE` so that `make.traits` returns an error if a process or its additional arguments (`process.args`) are not formatted correctly.
`make.traits` will send an error if the trait cannot be directly passed to `dads`.
However, in some specific cases (again, probably mainly for development and debugging) it could be useful to skip the tests using `test = FALSE`.

### 2023-04-05 inst/gitbook/03_make.modifiers.Rmd

The modifiers modify the core of the birth death process as implemented in `dads`.

@@

This contains a lot of information, much of it is actually not useful for using the modular aspects of `dads` at a high user level. I.e. unless you want to code very specific things, you won't need most of the information. The essential are these three functions in the elements named `"waiting$fun"`, `"selecting$fun"` and `"speciating$fun"`:

 1. `"waiting$fun"` is the branch length function (defined in details below) which returns a randomly drawn number from an exponential distribution with the rate of the number of taxa multiplied by the speciation and extinction rate: $n \times (\lambda + \mu)$ (here is uses the extra function `sum` for internal modularity reasons). **This function is responsible for the growth of the length/age of the tree**

```{r, eval = FALSE}
rexp(1, sum(lineage$n * (bd.params$speciation + bd.params$extinction)))
```
 With `lineage$n` being the number of lineages and `bd.params$speciation` and `bd.params$extinction` the speciation and extinction parameters (these specific terms are defined in details below).

 2. `"selecting$fun"` is the selection function (defined in details below) which, after the waiting time defined above returns a randomly selected lineage (a tip) among the existing ones. **This function is responsible for the branch selection**.

@@

 3. `"speciating$fun"` is the speciation function (defined in details below) which randomly draws a number (`runif(1)`) and depending on the speciation and extinction parameters makes the selected lineage (from 2.) speciate (`TRUE`) or go extinct (`FALSE`). **This function is responsible for the speciation or extinction of species**.

```{r, eval = FALSE}
runif(1) < (bd.params$speciation/(bd.params$speciation + bd.params$extinction)))
```

We will have a look at all these aspects below in more details which should make things clearer (but you can always refer to the default here for information).

@@

The first argument in `"modifiers"` is the branch length function (`branch.length`) this is the function that will be executed in `dads` to generate branch length.
Note that in the `dads` algorithm, branch length is not generated _directly_ but actually results of the waiting time.
In other words, the `branch.length` function just affects waiting time for all taxa present at any time in the simulation.
These taxa can then either go extinct (stopping the "growth" of it's branch length) or survive (continuing the "growth").

@@

Of course, the point of the modularity here is that you can provide your own function for generating branch length.
For example, we might be interested in what our tree would look like if we'd use a simple constant branch length generation (instead of randomly drawing it from an exponential distribution).
We can do so by declaring our own `branch.length` function and adding it to a `"modifiers"` object.

@@

```{r}
par(mfrow = c(1,2))
plot(default_tree,  main = "Default modifier")
plot(modified_tree, main = "Constant branch length\nmodifier")
```

It is of course possible to use more complex branch length modifiers that intakes different conditions and specific modification rather than simply always output a value of one.

### The allowed arguments {#allowarguments}

@@

  * `bd.params`: a named list containing `"numeric"` values that contains the birth death parameters (at least `"speciation"` and `"extinction"`);
  * `lineage`: a named list containing the lineage data (see below).
  * `trait.values`: a `"matrix"` containing `"numeric"` values with the trait names as column names and the lineages ID as row numbers (you can use it with the function `parent.traits` to access the trait of the previous node for example).
  * `modify.fun`: a `"list"` of named `"function"` (usually passed through `condition` and `modify`).

@@

In general, unless you know what you're doing, you can ignore most arguments for specific modifiers since they are handled automatically within the `dads` function.
Therefore any argument can be left undeclared or missing and is always handled internally. For example, if you did not declare `lineage$n` as a function argument but are using `lineage$n` in the function, `lineage$n` will be detected and treated as a current argument automatically as set accordingly within the birth death process (e.g. `lineage$n` will be set to the current number of taxa every iteration of the process).

For example, we can create a function that increases branch length proportional to the number of species "alive" at each time of the simulation in a discrete way. I.e. for discrete numbers of taxa, the branch length increases by jumps (ten fold) every 5 taxa:

@@

```{r}
par(mfrow = c(1,3))
plot(default_tree,    main = "Default modifier")
plot(modified_tree,   main = "Constant branch length\nmodifier")
plot(increasing_tree, main = "Increasing branch length\nmodifier (binned)")
```

@@

The `selection` function is used in the birth death process to know which lineage to select when running a speciation (or extinction!) event.
By default, this function randomly selects one taxon that is currently not extinct (using: `sample(1:lineage$n, 1))`.
Similarly as `branch.length` it is possible to modify this part of the birth death process.
For example, we could simply select always the last created lineage (to create a "ladder" or most asymmetric tree):

@@

> Note that here the function can only intake the [allowed arguments as described above](#allowarguments) (here `lineage$n`: the number of current living taxa).

We can then create a `"modifiers"` object the same way as before using this time the `selection` argument:

```{r}
## A modifier for selection
ladderised_modifier <- make.modifiers(selection = select.last)
## Generating a new tree with this modifier
set.seed(0)
ladderised_tree <- dads(bd.params = bd_params,
                        stop.rule = stop_rule,
                        modifiers = ladderised_modifier)
## Displaying the results
par(mfrow = c(1,2))
plot(default_tree,    main = "Default modifier")
plot(ladderised_tree, main = "Ladderising modifier")
```

@@

The third function that can be used to modify the birth death process is the `speciation` function.
This one is used during the birth death process to decide whether a lineages speciates (creating a node and two new lineages) or goes extinct (creating a tip).

@@

Creating a `"modifiers"` with a `speciation` function works the same way as for `branch.length` and `selection` but the function that will be used needs to output a logical value (see table [below](#summarymodifiers)).
Once the function is created simply input your function for speciation in the modifier and run the `dads` function with that modifier:

@@

par(mfrow = c(1,2))
plot(default_tree,  main = "Default modifier")
plot(modified_tree, main = "Random extinction\nmodifier")
```

> Note how loads of lineages end up going extinct even if the extinction parameter is set to 0!

And again, we can make some more advanced modifiers: for example, one where a tip always goes extinct if their ancestor has a negative trait value (here we will also introduce the utility function `parent.trait` that automatically selects the trait values of the parent of the current lineage.

```{r}
## A modifier for removing tips with negative values
bias.trait <- function(trait.values, lineage) {
    if(parent.traits(trait.values, lineage) < 0) {
        ## Go extinct!
        return(FALSE)
    } else {
        ## Speciate
        return(TRUE)
    }
}

@@

In the examples above, we have seen how to specify modifications to the birth death process (via `branch.length`, `selection` and `speciation`), however, one might note that these modifications are not dynamic.
In other words, throughout the process, the modifications remain constant (even if they are conditional).
It is however possible to code the `"modifiers"` so that they can be affected by `"events"` objects (see next chapter [on `events`](#makeevents)).

To do so, you can formally declare conditions (`condition`) and modifications (`modify`) as internal functions that can then be modified my an `"events"` object.
`condition` and `modify` are hard coded in the `branch.length` function that they concern, i.e. they are variables (functions) within the function.

@@

    ## Speciate?
    return(trigger_event < (bd.params$speciation/(bd.params$speciation + bd.params$extinction)))
}
```

It is possible with some specific condition to modify this trigger by providing a `condition` and `modify` function to the speciation function.
That is, if the `condition` is met, apply the `modify` function to the algorithm.
For example here we can edit the default speciation function to have a modification (`modify = double.the.trigger`: doubling the trigger value) happening half the time (`condition = half.the.time`):

```{r}
## A conditional function that triggers half the time
half.the.time <- function() return(sample(c(TRUE, FALSE), 1))
## A modification that doubles the value to trigger the event
double.the.trigger <- function(x) return(x*2)
## A conditional modifier
make.modifiers(speciation = speciation,
               condition  = half.the.time, 
               modify     = double.the.trigger)
```
Effectively, this will internally modify the speciation function as follows

@@

    ## Speciate?
    return(trigger_event < (bd.params$speciation/(bd.params$speciation + bd.params$extinction)))
}
```

These `condition` and `modify` functions can be applied to all the `modifiers` elements (`selection`, `branch.length` and `speciation`).
However, they are typically used in `events` that will modify... the `modifiers` ([see the `events` section](#makeevents))!

>>>>

The modifiers modify the core of the birth-death process as implemented in `dads`.

@@

This contains a lot of information, much of it is actually not useful for using the modular aspects of `dads` at a high user level, i.e. unless you want to code very specific things, you won't need most of the information. The essentials are these three functions in the elements named `"waiting$fun"`, `"selecting$fun"` and `"speciating$fun"`:

 1. `"waiting$fun"` is the branch length function (defined in detail below) which returns a randomly drawn number from an exponential distribution with the rate of the number of taxa multiplied by the speciation and extinction rate: $n \times (\lambda + \mu)$ (here is uses the extra function `sum` for internal modularity reasons). **This function is responsible for the growth of the length/age of the tree**

```{r, eval = FALSE}
rexp(1, sum(lineage$n * (bd.params$speciation + bd.params$extinction)))
```
 With `lineage$n` being the number of lineages and `bd.params$speciation` and `bd.params$extinction` the speciation and extinction parameters (these specific terms are defined in detail below).

 2. `"selecting$fun"` is the selection function (defined in detail below) which, after the waiting time defined above returns a randomly selected lineage (a tip) among the existing ones. **This function is responsible for the branch selection**.

@@

 3. `"speciating$fun"` is the speciation function (defined in detail below) which randomly draws a number (`runif(1)`) and depending on the speciation and extinction parameters makes the selected lineage (from 2.) speciate (`TRUE`) or go extinct (`FALSE`). **This function is responsible for the speciation or extinction of species**.

```{r, eval = FALSE}
runif(1) < (bd.params$speciation/(bd.params$speciation + bd.params$extinction))
```

We will have a look at all these aspects below in more detail which should make things clearer (but you can always refer to the default here for information).

@@

The first argument in `"modifiers"` is the branch length function (`branch.length`) this is the function that will be executed in `dads` to generate branch lengths.
Note that in the `dads` algorithm, branch length is not generated _directly_ but as the result of the waiting time.
In other words, the `branch.length` function just affects waiting time for all taxa present at any time in the simulation.
These taxa can then either go extinct (stopping the "growth" of its branch length) or survive (continuing the "growth").

@@

Of course, the point of the modularity here is that you can provide your own function for generating branch length.
For example, we might be interested in what our tree would look like if we use a simple constant branch length generation (instead of randomly drawing it from an exponential distribution).
We can do so by declaring our own `branch.length` function and adding it to a `"modifiers"` object.

@@

```{r}
par(mfrow = c(1,2))
plot(default_tree,  main = "Default modifier")
plot(modified_tree, main = "Constant branch length\nmodifier")
par(mfrow = c(1,1))
```

It is of course possible to use more complex branch length modifiers that take different conditions and specific modification rather than simply always outputing a value of one.

### The modifier arguments {#allowarguments}

@@

  * `bd.params`: a named list containing `"numeric"` values that contains the birth-death parameters (at least `"speciation"` and `"extinction"`);
  * `lineage`: a named list containing the lineage data (see below).
  * `trait.values`: a `"matrix"` containing `"numeric"` values with the trait names as column names and the lineage ID as row numbers (you can use it with the function `parent.traits` to access the trait of the previous node for example).
  * `modify.fun`: a `"list"` of named `"function"` (usually passed through `condition` and `modify`).

@@

In general, unless you know what you're doing, you can ignore most arguments for specific modifiers since they are handled automatically within the `dads` function.
Therefore any argument can be left undeclared or missing and is always handled internally. For example, if you did not declare `lineage$n` as a function argument but are using `lineage$n` in the function, `lineage$n` will be detected and treated as a current argument automatically as set accordingly within the birth-death process (e.g. `lineage$n` will be set to the current number of taxa every iteration of the process).

For example, we can create a function that increases branch length proportional to the number of species "alive" at each time of the simulation in a discrete way, i.e. for discrete numbers of taxa, the branch length increases by jumps (ten fold) every five taxa:

@@

```{r}
par(mfrow = c(1,3))
plot(default_tree,    main = "Default modifier")
plot(modified_tree,   main = "Constant branch length\nmodifier")
plot(increasing_tree, main = "Increasing branch length\nmodifier (binned)")
par(mfrow = c(1,1))
```

@@

The `selection` function is used in the birth-death process to know which lineage to select when running a speciation (or extinction!) event.
By default, this function randomly selects one taxon that is currently not extinct (using: `sample(1:lineage$n, 1))`.
Similarly to `branch.length`, it is possible to modify this part of the birth-death process.
For example, we could simply select the last created lineage to create a "ladder" or most asymmetric tree:

@@

> Note that here the function can only take the [allowed arguments as described above](#allowarguments) (here `lineage$n`: the number of current living taxa).

We can then create a `"modifiers"` object the same way as before this time using the `selection` argument:

```{r}
## A modifier for selection
ladderised_modifier <- make.modifiers(selection = select.last)
## Generating a new tree with this modifier
set.seed(0)
ladderised_tree <- dads(bd.params = bd_params,
                        stop.rule = stop_rule,
                        modifiers = ladderised_modifier)
## Displaying the results
par(mfrow = c(1,2))
plot(default_tree,    main = "Default modifier")
plot(ladderised_tree, main = "Ladderising modifier")
par(mfrow = c(1,1))
```

@@

The third function that can be used to modify the birth-death process is the `speciation` function.
This one is used during the birth-death process to decide whether a lineage speciates (creating a node and two new lineages) or goes extinct (creating a tip).

@@

Creating `"modifiers"` with a `speciation` function works the same way as for `branch.length` and `selection` but the function that will be used needs to output a logical value (see table [below](#summarymodifiers)).
Once the function is created simply input your function for speciation in the modifier and run the `dads` function with that modifier:

@@

par(mfrow = c(1,2))
plot(default_tree,  main = "Default modifier")
plot(modified_tree, main = "Random extinction\nmodifier")
par(mfrow = c(1,1))
```

> Note how loads of lineages go extinct even if the extinction parameter is set to 0!

And again, we can make some more advanced modifiers: for example, one where a tip always goes extinct if their ancestor has a negative trait value. Here we will also introduce the utility function `parent.trait` that automatically selects the trait values of the parent of the current lineage.

```{r}
## A modifier for removing tips with negative values
bias.trait <- function(trait.values, lineage) {
    if(parent.traits(trait.values, lineage) < 0) {
        ## Go extinct!
        return(FALSE)
    } else {
        ## Speciate!
        return(TRUE)
    }
}

@@

In the examples above, we have seen how to specify modifications to the birth-death process (via `branch.length`, `selection` and `speciation`), however, these modifications are not dynamic.
In other words, throughout the process, the modifications remain constant (even if they are conditional).
It is however, possible to code the `"modifiers"` so that they can be affected by `"events"` objects (see next chapter [on `events`](#makeevents)).

To do so, you can formally declare conditions (`condition`) and modifications (`modify`) as internal functions that can then be modified by an `"events"` object.
`condition` and `modify` are hard coded in the `branch.length` function that they concern, i.e. they are variables (functions) within the function.

@@

    ## Speciate?
    return(trigger_event < (bd.params$speciation/
                              (bd.params$speciation + bd.params$extinction)))
}
```

It is possible with some specific conditions to modify this trigger by providing a `condition` and `modify` function to the speciation function.
That is, if the `condition` is met, apply the `modify` function to the algorithm.
For example here we can edit the default speciation function to have a modification (`modify = double.the.trigger`: doubling the trigger value) happening half the time (`condition = half.the.time`):

```{r}
## A conditional function that triggers half the time
half.the.time <- function() return(sample(c(TRUE, FALSE), 1))
## A modification that doubles the value to trigger the event
double.the.trigger <- function(x) return(x*2)
## A conditional modifier
make.modifiers(speciation = speciation,
               condition  = half.the.time, 
               modify = double.the.trigger)
```
Effectively, this will internally modify the speciation function as follows:

@@

    ## Speciate?
    return(trigger_event < (bd.params$speciation/
                              (bd.params$speciation + bd.params$extinction)))
}
```

These `condition` and `modify` functions can be applied to all the `modifiers` elements (`selection`, `branch.length` and `speciation`).
However, they are typically used in `events` that will modify the `modifiers` ([see the `events` section](#makeevents))!
